Approximately Exact Calculations for Linear Mixed Models

### 2015-05-12 ms.tex
added Bray as author.  Added appendix

\usepackage{amsmath,amsfonts}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{caption,subcaption}
%\usepackage{color}
%\usepackage{etoolbox}
\usepackage{float}
\usepackage{fullpage}
%\usepackage{mathtools}
\usepackage{natbib}
\usepackage{paralist}
%\usepackage{setspace}

@@

\title{Approximately Exact Calculations for Linear Mixed Models}
\author[1]{Michael Lavine}
\author[2]{James Hodges}
\affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
\affil[2]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}

@@

\cite{hodges:2013} also unifies and generalizes \cite{reich_hodges:2008} and \cite{welham_thompson:2009} to show that in our special case, and a few others, $\logRLssorig$ can be expressed as
\begin{equation}
\label{eq:reexpress}
  \logRLssorig = B - \frac{n_e}{2}\log(\sigesq) - \frac{y^t \Gamma_c \Gamma^t_c y}{2\sigesq} 
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right]
\end{equation}
where $B$ is an unimportant known constant; $n_e$ is $n$ minus the dimension of the space spanned by the columns of $[X|Z]$; $\Gamma_c$ is $n \times n_e$ and spans the space orthogonal to $[X|Z]$ (so $y^t \Gamma_c \Gamma^t_c y$ is the residual sum of squares); $s_z$ is the dimension of the space spanned by the columns of $Z$ not already in the span of the columns of $X$; and the $\{a_j\}$ and $\{\hat v_j\}$ are known constants with all $a_j>0$.  Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.  Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\abss$ of the two unknowns.

@@

\section{An Algorithm}
With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
\begin{enumerate}
\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
	\begin{enumerate}[(a)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		the algorithm's loop.
	\item We will not carefully analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
	control the running of the algorithm.  Their meaning may become more
	clear as we explain the algorithm.
\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
	and \eqref{eq:rll} for definitions --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
	\dots, s_z, v, h$.
\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
	lower bound on $\logRL(\sshat)$.
\item While the active list is not empty or we have not reached \maxit\
	iterations
	\begin{enumerate}
	\item For each active box $b$:
		\begin{enumerate}
		\item if $U^b < L-M$ move $b$ to the inactive list.
		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
		\item if the horizontal extent of $b$ is less than $\delta_e$,
			move $b$ to the inactive list
		\item if the vertical extent of $b$ is less than $\delta_s$,
			move $b$ to the inactive list
		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
			Remove $b$ and add the subboxes to the active list.
		\end{enumerate}
	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
		lower bound on $\logRL(\sshat)$.
	\end{enumerate}
\item Return the active and inactive lists.
\end{enumerate}

@@

As written, our algorithm moves a box $b$ to the inactive list if
\begin{enumerate}[(a)]
\item $U^b < L-M$ or
\item $U^b - L^b < \epsilon$ or
\item either the vertical or horizontal extent of $b$ is sufficiently small.
\end{enumerate}
But one could construct more elaborate rules.  One appealing example is to apply criteria (a) and (c) if $U^b \le L-\epsilon_2$ and apply criterion (b) if $U^b > L-\epsilon_2$.  Other rules are possible, too.  We don't elaborate here in order to concentrate on the main ideas.

@@

>>>>

\usepackage{amsmath,amsfonts}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{caption,subcaption}
%\usepackage{color}
%\usepackage{etoolbox}
\usepackage{float}
\usepackage{fullpage}
%\usepackage{mathtools}
\usepackage{natbib}
\usepackage{enumitem}
%\usepackage{appendix}
%\usepackage{setspace}

@@

\title{Approximately Exact Calculations for Linear Mixed Models}
\author[1]{Michael Lavine}
\author[1,2]{Andrew Bray}
\author[3]{James Hodges}
\affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
\affil[2]{Mount Holyoke College}
\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}

@@

\cite{hodges:2013} also unifies and generalizes \cite{reich_hodges:2008} and \cite{welham_thompson:2009} to show that in our special case, and a few others, $\logRLssorig$ can be expressed as
\begin{equation}
\label{eq:reexpress}
  \logRLssorig = B - \frac{n_e}{2}\log(\sigesq) - \frac{y^t \Gamma_c \Gamma^t_c y}{2\sigesq} 
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right]
\end{equation}
where
\begin{enumerate}[label=(\arabic*)]
  \item $B$ is an unimportant known constant;
  \item $n_e$ is $n$ minus the dimension of the space spanned by the columns of $[X|Z]$;
  \item $\Gamma_c$ is $n \times n_e$ and spans the space orthogonal to $[X|Z]$
    (so $y^t \Gamma_c \Gamma^t_c y$ is the residual sum of squares);
  \item $s_z$ is the dimension of the space spanned by the columns of $Z$ not already
    in the span of the columns of $X$; and
  \item the $\{a_j\}$ and $\{\hat v_j\}$ are known constants whose derivation is in the Appendix.  All $a_j>0$.
\end{enumerate}
Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.  Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\abss$ of the two unknowns.

@@

\section{An Algorithm}
With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
\begin{enumerate}
\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
	\begin{enumerate}[label=(\alph*)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		the algorithm's loop.
	\item We will not carefully analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
	control the running of the algorithm.  Their meaning may become more
	clear as we explain the algorithm.
\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
	and \eqref{eq:rll} for definitions --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
	\dots, s_z, v, h$.
\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
	lower bound on $\logRL(\sshat)$.
\item While the active list is not empty or we have not reached \maxit\
	iterations
	\begin{enumerate}
	\item For each active box $b$:
		\begin{enumerate}
		\item if $U^b < L-M$ move $b$ to the inactive list.
		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
		\item if the horizontal extent of $b$ is less than $\delta_e$,
			move $b$ to the inactive list
		\item if the vertical extent of $b$ is less than $\delta_s$,
			move $b$ to the inactive list
		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
			Remove $b$ and add the subboxes to the active list.
		\end{enumerate}
	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
		lower bound on $\logRL(\sshat)$.
	\end{enumerate}
\item Return the active and inactive lists.
\end{enumerate}

@@

As written, our algorithm moves a box $b$ to the inactive list if
\begin{enumerate}[label=(\alph*)]
\item $U^b < L-M$ or
\item $U^b - L^b < \epsilon$ or
\item either the vertical or horizontal extent of $b$ is sufficiently small.
\end{enumerate}
But one could construct more elaborate rules.  One appealing example is to apply criteria (a) and (c) if $U^b \le L-\epsilon_2$ and apply criterion (b) if $U^b > L-\epsilon_2$.  Other rules are possible, too.  We don't elaborate here in order to concentrate on the main ideas.

@@

\section{Appendix}
\paragraph{Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}}  Our derivation follows \cite{hodges:2013},
which contains more details.  There are three steps.
\begin{enumerate}
  \item \textbf{Make the covariance matrices proportional to the identity.}  If $\Sigma_e$ is not the identity matrix,
    transform the data to $\Sigma_e^{-.5}y$.  The transformed data, which we shall still call $y$, has covariance
    proportional to the identity.  Similarly, if $\Sigma_s$ is not the identity matrix, re-parameterize the random
    effects to $\Sigma_s^{-.5}u$.  The re-parameterized random effects, which we shall still call $u$, have
    covariance proportional to the identity.
  \item \textbf{If the column spaces of $X$ and $Z$ have a non-trivial intersection, transform them.}
    Let $s_X=\text{rank}(X)$  and $s_Z=\text{rank}(X|Z) - s_X$.  Let $\Gamma_X$ be an $n \times s_X$ matrix whose
    columns are an orthonormal basis for the column space of $X$.  Let $\Gamma_Z$ be an $n \times s_Z$ matrix
    such that the columns of $\left[\Gamma_X | \Gamma_Z \right]$ are an orthonormal basis for the column space
    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times n-s_X-s_Z$ matrix such that the columns of
    $\left[\Gamma_X | \Gamma_Z | \Gamma_c\right]$ are an orthonormal basis for $\mathbb{R}^n$.  Define the matrix
    \begin{equation*}
      M = \begin{bmatrix}
               M_{XX} & M_{XZ}\\
               0 & M_{ZZ}
             \end{bmatrix}
    \end{equation*}
    by $[X|Z] = [\Gamma_X|\Gamma_Z]M$ where $M_{XX}$ is $s_X \times p$ and $M_{XZ}$ is $s_Z \times q$.
    $\Gamma_X$ and $\Gamma_Z$ are transformed versions of $X$ and $Z$ that have non-overlapping
    column spaces.
  \item \textbf{Re-parameterize and diagonalize.}  Let $M_{ZZ}$ have the singular value decomposition
    $PA^{.5}L^t$.  Now the linear mixed model \eqref{eq:lmm} can be written as
    \begin{equation*}
      \begin{split}
        y &= \begin{bmatrix} X Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X  \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X \Gamma_ZP \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
      \end{split}
    \end{equation*}
    where $\beta^* = M_{XX}\beta + M_{XZ}u$ and $v = A^{.5}L^t u$.  $\beta^*$ contains the re-parametrized
    fixed effects while $v$ contains the re-parametrized random effects.  The corresponding design matrices
    $\Gamma_X$ and $\Gamma_Z P$ are orthogonal to each other.
\end{enumerate}
Finally, the $\{a_j\}$ in \eqref{eq:reexpress} are the diagonal elements of $A$, all of which are strictly positive, and the $\{\hat v_j\}$ in \eqref{eq:reexpress} are given by $\hat v = (\hat v_1, \dots, \hat v_{s_Z})^t = P^t \Gamma_Z^t y$.
